---
title: "Exploratory Analysis"
author: "Grant Foster"
date: "3/29/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(PVR)
library(ape)
library(phytools)
library(phytools)
library(igraph)
library(magrittr)
```
# Table of Contents:
## 1. Galler Data Exploration
## 2. Subsample Empirical Trees Randomly & Perform Eigendecomposition
## 3.  Generating Random trees


# 1. Galler Data Exploration
```{r}
galltree <- ape::read.nexus(file="../Data.nosync/Nyman_Phylogeny.nex")
galltree <- phytools::force.ultrametric(galltree)

litter <- PVR::PVRdecomp(phy=galltree, type="nexus")

EigVec <- data.frame(litter@Eigen$vectors)
EigVal <- data.frame(litter@Eigen$values)
EigVal$litter.Eigen.values <- EigVal$litter.Eigen.values/(sum(EigVal$litter.Eigen.values))

phytools::plotTree.wBars(tree=galltree, setNames(EigVec$c1+abs(min(EigVec$c1)), galltree$tip.label))
phytools::plotTree.wBars(tree=galltree, setNames(EigVec$c2+abs(min(EigVec$c2)), galltree$tip.label))
```


```{r}
sharks_full <- ape::read.nexus(file="../Data.nosync/Stein2018SharkTree/output.nex")
shark_tree <- sharks_full[[1]]

litter <- PVR::PVRdecomp(phy=shark_tree, type="nexus")

EigVec <- data.frame(litter@Eigen$vectors)
EigVal <- data.frame(litter@Eigen$values)
EigVal$litter.Eigen.values <- EigVal$litter.Eigen.values/(sum(EigVal$litter.Eigen.values))

phytools::plotTree.wBars(tree=shark_tree, setNames(EigVec$c1+abs(min(EigVec$c1)), shark_tree$tip.label)) #This plot looks terrible but it validates our method
```

I got this from Upham Paper Dryad: https://datadryad.org/stash/dataset/doi:10.5061/dryad.tb03d03

The downloaded fileset (which is huge) has this file under "Data_S3_globalRAxML_files/RAxML_bipartitions.result_FIN4_raw_rooted_wBoots_4098mam1out_OK.newick"


Running this decomposition many times is a lot to ask of my little laptop. I'll get things set up and then try it on a bigger machine.
```{r}
mammals <- phytools::read.newick(file="../Data.nosync/RAxML_bipartitions.result_FIN4_raw_rooted_wBoots_4098mam1out_OK.newick")

mammals$node.label <- NULL #Igraph doesn't like the node labels given is this default tree, so we just remove them
```

litter <- PVR::PVRdecomp(phy=mammals)

EigVec <- data.frame(litter@Eigen$vectors)
EigVal <- data.frame(litter@Eigen$values)
EigVal$litter.Eigen.values <- EigVal$litter.Eigen.values/(sum(EigVal$litter.Eigen.values))

phytools::plotTree.wBars(tree=mammals, setNames(EigVec$c1+abs(min(EigVec$c1)), mammals$tip.label))



This is our tiny baby tree; it only has 15 nodes!
```{r}
spagTree <-  ape::read.nexus(file="../Data.nosync/Piatkowski&Shaw_2019_MCC_Tree.nex")
```

## 2. Subsample Empirical Trees Randomly & Perform Eigendecomposition
For multiple trees
```{r}
analyze <- function(treesList, treeNames, nreps){
  output <- NULL
  ntrees <- length(treesList)
  for(i in 1:nreps){
    for(j in 1:ntrees){
      tree <- treesList[[j]]
      #Choose a random set of nodes to remove
      numDrops <- sample(1:(length(tree$tip.label)-10), size=1)
      Intdrops <- sample(1:length(tree$tip.label), size=numDrops, replace = FALSE)
      
      #Get the descendants of those nodes and remove them
      randPruned <- ape::drop.tip(tree, tree$tip.label[Intdrops])
      
      randGraph <- igraph::as.igraph(randPruned)
      
      #Run a phylodecomp on that tree
      randDecomp <- PVR::PVRdecomp(phy=randPruned, type="nexus")
      
      randDF <- data.frame("rep"= i, 
                 "vectorID"=names(((randDecomp@Eigen$values)/sum(randDecomp@Eigen$values))[1:10]), 
                 "PropVar"=((randDecomp@Eigen$values)/sum(randDecomp@Eigen$values))[1:10],
                 "Ntips"=length(randPruned$tip.label), 
                 "infomapModularity"=igraph::infomap.community(randGraph)$modularity, 
                 "tipsDiscarded"=toString(sort(Intdrops)),
                 "tree"=treeNames[j], 
                 row.names = NULL
                 )
      #Save outputs
    output <- rbind(output, randDF)
    }
  }
  return(output)
}
```

Run this 

```{r}
#FastResults <- analyze(treesList = list(shark_tree, spagTree, galltree), treeNames = c("sharks", "spag", "gallers"), nreps=1000)
#save(FastResults, file="FastResults.RData")
load(file="FastResults.RData")
```


```{r}
p1 <- FastResults %>% dplyr::filter(., vectorID=="c1") %>%
  ggplot(data=., aes(x=log(Ntips), y=PropVar, color=as.factor(tree)))+geom_point(alpha=0.33)+
  xlab("Log Number of Species in Tree")+
  ylab("Proportion of Variance: 1st Eigenvector")+theme_classic()+
  theme(legend.position = c(0.7, 0.7))+guides(col=guide_legend("Phlogenetic Tree"))

p2 <- FastResults %>% dplyr::filter(., vectorID=="c2") %>%
  ggplot(data=., aes(x=log(Ntips), y=PropVar, color=as.factor(tree)))+geom_point(alpha=0.33)+
  xlab("Log Number of Species in Tree")+
  ylab("Proportion of Variance: 2nd Eigenvector")+
  theme_classic()+
  theme(legend.position ="none")

p3 <- FastResults %>% dplyr::filter(., vectorID=="c10") %>%
  ggplot(data=., aes(x=log(Ntips), y=PropVar, color=as.factor(tree)))+geom_point(alpha=0.33)+
  xlab("Log Number of Species in Tree")+
  ylab("Proportion of Variance: 10th Eigenvector")+
  theme_classic()+
  theme(legend.position ="none")

p4 <- FastResults %>% group_by(., rep, tree) %>% mutate(eigenSD=sd(PropVar, na.rm=TRUE)) %>%
  dplyr::select(., Ntips, eigenSD, tree) %>% unique() %>%
  ggplot(data=., aes(x=log(Ntips), y=eigenSD, color=as.factor(tree)))+
  geom_point(alpha=0.33)+
  xlab("Log Number of Species in Tree")+
  ylab("SD of First 10 Eigenvalues' Proportions Variance")+
  theme_classic()+
  theme(legend.position ="none")

#pdf(file="../Figures/VariancexNtips_Fast.pdf", width=12, height=11.5)
gridExtra::grid.arrange(p1, p2, p3, p4, nrow = 2, bottom="Figure1: The relationship between the log number of species in a subtree and the proportion of variance 
                        explained by the 1st, 2nd, or 10th largest eigenvalues (panels a-c), as well as the standard deviation of the variance
                        explained by first 10 eigenvalues. Each point is a tree constructed by randomly trimming tips from a larger complete tree.
                        Point color corresponds to the complete tree a given subtree is constructed from; all 1190 known sharks(Stein2018), a group
                        of 72 sawflies within subtribe Euurina (Kopelke), and a tree of 15 peat mosses in the genus Spaghnum (Piatkowski & Shaw, 2019)")
```



```{r}

mod1 <- FastResults %>% dplyr::filter(., vectorID=="c1") %>%
  ggplot(data=., aes(x=infomapModularity, y=PropVar, color=tree))+geom_point()+
  xlab("Modularity")+
  ylab("Proportion of Var Explained by 1st Eigenvector")+theme_classic()+
  theme(legend.position = c(0.7, 0.7))+guides(col=guide_legend("Phlogenetic Tree"))

mod2 <- FastResults %>% dplyr::filter(., vectorID=="c1") %>% dplyr::filter(., infomapModularity > 0) %>%
  ggplot(data=., aes(x=infomapModularity, y=PropVar, color=tree))+geom_point()+
  xlab("Non-Zero Modularity")+
  ylab("Proportion of Var Explained by 1st Eigenvector")+theme_classic()+
  theme(legend.position = c(0.2, 0.2))+guides(col=guide_legend("Phlogenetic Tree"))

#pdf(file="../Figures/ModularityxEig1.pdf", width=12, height=8.5)
gridExtra::grid.arrange(mod1, mod2, nrow = 1,
                        bottom="Figure2: The relationship between graph Modularity and the variance explained by the first eigenvalue. Points represent an individual
                        subtree, constructed as described above. Modularity is calculated using the algorthithm infoMap; clustering at 0 occurs due to the algorithm
                        being unwilling to assign membership below a certain modularity threshold. The rightmost panel is the same data, but with the 0-modularity
                        values removed.")

```

###START HERE!!!!!
### Last thing done: messed around with taking distribution summary stats (1st 3 moments) of distance matrix entries to see if that holds more variance. It does, but doesn't capture clustering (except maybe indirectly). Either do the decomp adn see just for giggles, or think about another clustering metric next (Peixoto?)

# 3. Generating Random trees
```{r}
randTrees <- ape::rmtree(40, rooted=TRUE, N=100)

dmat <- list()
output <- NULL
for(i in 1:100){
  #ultree <- force.ultrametric(randTrees[[i]])
  ultree <- randTrees[[i]]
  randGraph <- igraph::as.igraph(ultree)
  value <- igraph::infomap.community(randGraph)$modularity
  dmat[[i]] <- ape::cophenetic.phylo(ultree)
  output <- c(output, value)
}

hist(dmat[[1]])
hist(dmat[[2]])
hist(dmat[[3]])
hist(dmat[[4]])
hist(dmat[[5]])
hist(dmat[[6]])
hist(dmat[[7]])
hist(dmat[[8]])


dmat[[1]]


x <- MASS::fitdistr(dmat[[1]], densfun = "normal", method = "mle")
as.vector(dmat[[1]])
skewness
fit.weibull <- MASS::fitdistr(dmat[[1]], densfun = "weibull", method = "mle", lower = c(0, 0))

skew <- NULL
ex_kurt <- NULL
avg <- NULL
for(i in 1:100){
  avg <- c(avg, mean(as.vector(dmat[[i]])))
  skew <- c(skew, moments::skewness(as.vector(dmat[[i]])))
  ex_kurt <- c(ex_kurt, moments::kurtosis(as.vector(dmat[[i]]))-3)
}
hist(avg)
hist(skew)
hist(ex_kurt)
```


```{r}
randGraph <- igraph::as.igraph(ult_Tree)

analyze <- function(treesList, treeNames, nreps){
  output <- NULL
  ntrees <- length(treesList)
  for(i in 1:nreps){
    for(j in 1:ntrees){
      tree <- treesList[[j]]
      #Choose a random set of nodes to remove
      numDrops <- sample(1:(length(tree$tip.label)-10), size=1)
      Intdrops <- sample(1:length(tree$tip.label), size=numDrops, replace = FALSE)
      
      #Get the descendants of those nodes and remove them
      randPruned <- ape::drop.tip(tree, tree$tip.label[Intdrops])
      
      randGraph <- igraph::as.igraph(randPruned)
      
      #Run a phylodecomp on that tree
      randDecomp <- PVR::PVRdecomp(phy=randPruned, type="nexus")
      
      randDF <- data.frame("rep"= i, 
                 "vectorID"=names(((randDecomp@Eigen$values)/sum(randDecomp@Eigen$values))[1:10]), 
                 "PropVar"=((randDecomp@Eigen$values)/sum(randDecomp@Eigen$values))[1:10],
                 "Ntips"=length(randPruned$tip.label), 
                 "infomapModularity"=igraph::infomap.community(randGraph)$modularity, 
                 "tipsDiscarded"=toString(sort(Intdrops)),
                 "tree"=treeNames[j], 
                 row.names = NULL
                 )
      #Save outputs
    output <- rbind(output, randDF)
    }
  }
  return(output)
}

randtree <- ape::rtree(n=40, rooted=TRUE)
randtree$edge
class(randtree)
plot(randtree)
ult_Tree <- force.ultrametric(randtree)
plot(ult_Tree)

ult_graph <- igraph::as.igraph(ult_Tree)
igraph::infomap.community(ult_graph)$modularity
```
How modular can I make a tree?

```{r}
BatsnMarsupsTrees <- read.nexus(file="../Data.nosync/BatsnMarsupTrees.nex")
test <- force.ultrametric(BatsnMarsupsTrees[[1]])
ult_graph <- igraph::as.igraph(test)

igraph::infomap.community(ult_graph)$modularity


dmat <- ape::cophenetic.phylo(test)
comm <- as.data.frame.matrix(table(test))

test_mpd <- ses.mpd(comm, dmatrix, null.model = "independentswap", runs = 99, iterations = 1000, abundance.weighted = FALSE) 
```

